A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-12-17 main.tex
Edits to Section 2.1 and 2.2

Population structure is defined by the existence of allele frequency differences that distinguish subpopulations and is driven by the combined effects of evolutionary processes such as genetic drift, migration, demographic history, and natural selection \citep{gibson2015primer, tibayrenc2017genetics}. Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Ancient relatedness describes the presence of a common ancestor many generations previously. The presence of distinct ancestry groups with different allele frequencies in a sample is known as \textbf{population stratification}. Recent relatedness describes the sharing of a common ancestor only several generations previously. Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. In the absence of known familial relationships, recent relatedness is referred to as \textbf{cryptic relatedness.}

@@

In this review we focus on methods to account for unobserved relatedness such as population stratification and cryptic relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations when population structure is associated with differences in both allele frequency and the trait or disease of interest \citep{gibson2015primer}. This phenomenon is commonly described as \textbf{confounding due to population stratification}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure in and of itself does not confound the genotype-phenotype relationship. While population structure may directly affect observed allele frequencies, there must also exist some non-genetic mechanism by which population stratification affects the phenotype, namely, the environment \citep{barton2019population, vilhjalmsson2012nature}.

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but has no direct effect on lung cancer. Also suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would likely find SNP $X$ to be significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in diagram \eqref{diag:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question.

@@

An additional source of confounding in genetic studies that we will briefly mention is that of confounding due to genetic background. This occurs in univariate association methods when a non-causal SNP is tested for association with a trait of interest, and that SNP is correlated with a causal SNP. As \citet{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. LD pruning is commonly employed to reduce confounding due to genetic background in univariate association testing. However, we will focus on methods which address this via genomewide multivariate analysis.

>>>>

Population structure is defined by the existence of allele frequency differences that distinguish subpopulations and is driven by the combined effects of evolutionary processes such as genetic drift, migration, demographic history, and natural selection \citep{gibson2015primer, tibayrenc2017genetics}. Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Ancient relatedness describes the presence of a common ancestor many generations previously. The presence of distinct ancestry groups with different allele frequencies in a sample is known as \textbf{population stratification}. Recent relatedness describes the sharing of a common ancestor only a few generations previous. Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. In the absence of known familial relationships, recent relatedness is referred to as \textbf{cryptic relatedness.}

@@

In this review we focus on methods to account for unobserved dependence between samples, whether arising from population stratification or cryptic relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations when population structure is associated with differences in both allele frequency and the trait or disease of interest \citep{gibson2015primer}. This phenomenon is commonly described as \textbf{confounding due to population stratification}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure in and of itself does not confound the genotype-phenotype relationship. Rather, there must also exist some non-genetic mechanism by which population stratification affects the phenotype, namely, the environment \citep{barton2019population, vilhjalmsson2012nature}.

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is more common in subpopulation B compared to subpopulation A, but has no direct effect on lung cancer. Also suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would likely find SNP $X$ to be significantly associated with lung cancer. This spurious association is does not arise from the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be associated with the phenotype.

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in diagram \eqref{diag:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction.
%Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related.
In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality, only that they are now correlated. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question.

@@

An additional source of confounding in genetic studies that we will briefly mention is that of confounding due to genetic background. This occurs in univariate association methods when a non-causal SNP is tested for association with a trait of interest, and that SNP is correlated with a causal SNP. As \citet{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. LD pruning is commonly employed to reduce confounding due to genetic background in univariate association testing. Here, we focus on methods which address this via genomewide multivariate analysis.
